Towards atomistic resolution structure of phosphatidylcholine headgroup and glycerol backbone at different ambient conditions\footnote{Publication about results presented in the NMRlipids project}

### 2015-05-15 HGmodelMANUSCRIPT/HGmodel_ACStemplate.tex
Methods: Order parameters from experiments.

\subsection{Order parameters from experiments}\label{expORDp}
The order parameter of a hydrocarbon C--H vector is defined as 
\begin{equation}\label{orderP}
S_{{\rm CH}}=\frac{1}{2}\langle 3 \cos^2 \theta-1 \rangle,
\end{equation} 
where the angle brackets denote an ensemble average over the sampled conformations, and $\theta$ is the angle between the C--H bond and the membrane normal.
The absolute values of order parameters can be measured by detecting quadrupolar splitting with $^2$H NMR~\cite{seelig77c} or by detecting dipolar 
splitting with $^1$H-$^{13}$C NMR~\cite{hong95a,gross97,dvinskikh05a,ferreira13}. The measurements are based on
different physical interactions and also the connection between order parameters and quadrupolar or dipolar splitting
are different. The absolute values of order parameters from the measured quadrupolar splitting $\Delta \nu_Q$ ($^2$H NMR) are calculated using 
the equation $|S_{{\rm CD}}|=\frac{4}{3} \frac{e^2qQ}{h} \Delta \nu_Q$, where the value for the static quadrupole
splitting constant is estimated from various experiments to be 170~kHz leading to a numerical relation $|S_{{\rm CD}}|=0.00784 \times \Delta \nu_Q$~\cite{seelig77c}. 
The absolute values of order parameters from the effective dipolar coupling $d_\mathrm{CH}$ ($^1$H-$^{13}$C NMR) are calculated using equation
$|S_{{\rm CH}}|=\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c} d_\mathrm{CH}$, where
values between 20.2--22.7 kHz are used for $\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c}$,
depending on the original authors~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
The effective dipolar coupling $d_\mathrm{CH}$ is related to the measured dipolar splitting $\Delta \nu_{CH}$ 
through scaling factor which depends on the pulse sequence used in the $^1$H-$^{13}$C NMR experiment~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
It is important to note that the order parameters measured with different techniques based on different physical interactions are in good agreement
with each other (see Results and Discussion), indicating very high quantitative accuracy of the measurements.
For a more detailed discussion see Ref.~\citenum{accuracyPOST}.

The absolute values of order parameters are accessible with both $^2$H NMR and $^1$H-$^{13}$C NMR techniques. 
However, only $^1$H-$^{13}$C NMR techniques allow also the measurement of the sign of the order parameter~\cite{hong95a,hong95b,gross97}. 
The measured sign is negative for almost all the carbons discussed in this work as only $\alpha$ is positive~\cite{hong95a,hong95b,gross97}. 
For more detailed discussion about the sign measurements of the order parameters see Ref.~\citenum{signpost}.

For most CH$_2$ segments in fluid phosholipid bilayer the order parameters are equal for both hydrogens attached to the same carbon.
However, in some cases (e.g. g$_1$, g$_3$ and  C$_2$ carbon in the \textit{sn}-2 chain) the order parameters are not equal and this 
can be observed with both $^2$H NMR and $^1$H-$^{13}$C NMR techniques. In the present work we call the phenomena of unequal order parameters 
for hydrogens attached to the same carbon as {\it forking} to avoid confusion with dipolar and quadrupolar splitting in NMR terminology. 
Forking has been studied in detail with $^2$H NMR techniques by deuterating the R or S position in CH$_2$ segment, and
the studies show that the forking arises from differently sampled orientations of the two C--H bonds, not from two 
separate populations of lipid conformations~\cite{engel81,gally81}.

>>>>

\subsection{Order parameters from experiments}\label{expORDp}
The order parameter of a hydrocarbon C--H vector is defined as 
\begin{equation}\label{orderP}
S_{{\rm CH}}=\frac{1}{2}\langle 3 \cos^2 \theta-1 \rangle,
\end{equation} 
where the angle brackets denote an ensemble average over the sampled conformations, and $\theta$ is the angle between the C--H bond and the membrane normal.
The absolute values of order parameters can be measured by detecting quadrupolar splitting with $^2$H NMR~\cite{seelig77c} or by detecting dipolar 
splitting with $^1$H-$^{13}$C NMR~\cite{hong95a,gross97,dvinskikh05a,ferreira13}. The measurements are based on
different physical interactions, and also the connections between order parameter and quadrupolar or dipolar splitting
are different. The absolute values of order parameters from the measured quadrupolar splitting $\Delta \nu_Q$ ($^2$H NMR) are calculated using 
the equation $|S_{{\rm CD}}|=\frac{4}{3} \frac{e^2qQ}{h} \Delta \nu_Q$, where the value for the static quadrupole
splitting constant is estimated from various experiments to be 170~kHz leading to a numerical relation $|S_{{\rm CD}}|=0.00784 \times \Delta \nu_Q$~\cite{seelig77c}. 
The absolute values of order parameters from the effective dipolar coupling $d_\mathrm{CH}$ ($^1$H-$^{13}$C NMR) are calculated using the equation
$|S_{{\rm CH}}|=\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c} d_\mathrm{CH}$, where
values between 20.2--22.7~kHz are used for $\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c}$,
depending on the original authors~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
The effective dipolar coupling $d_\mathrm{CH}$ is related to the measured dipolar splitting $\Delta \nu_\mathrm{CH}$ 
through a scaling factor that depends on the pulse sequence used in the $^1$H-$^{13}$C NMR experiment~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
It is important to note that the order parameters measured with different techniques based on different physical interactions are in good agreement
with each other (see Results and Discussion), indicating very high quantitative accuracy of the measurements.
For a more detailed discussion, see Ref.~\citenum{accuracyPOST}.

The absolute values of order parameters are accessible with both $^2$H NMR and $^1$H-$^{13}$C NMR techniques. 
However, only $^1$H-$^{13}$C NMR techniques allow also the measurement of the sign of the order parameter~\cite{hong95a,hong95b,gross97}. 
The measured sign is negative for almost all the carbons discussed in this work, only the choline $\alpha$ is positive~\cite{hong95a,hong95b,gross97}. 
For more detailed discussion about the sign measurements of the order parameters, see Ref.~\citenum{signpost}.

For most CH$_2$ segments in a fluid phospholipid bilayer, the order parameters of both hydrogens are equal.
However, in some cases (e.g., g$_1$, g$_3$, and  C$_2$ carbon in the \textit{sn}-2 chain) the two order parameters are not equal;
this  can be observed with both $^2$H NMR and $^1$H-$^{13}$C NMR techniques. In the present work,
to avoid confusion with the dipolar and quadrupolar splittings in NMR terminology,
we call the phenomenon of unequal order parameters for hydrogens attached to the same carbon {\it forking}. 
Forking has been studied in detail with $^2$H NMR techniques by deuterating the R or S position in CH$_2$ segment, and
the studies show that it arises from differently sampled orientations of the two C--H bonds, not from two 
separate populations of lipid conformations~\cite{engel81,gally81}.
